# Central Limit Theorem {#TCL}

The law of large numbers (LLN) and the central limit theorem (CLT) answer two different questions about sample averages. The LLN explains why a sample mean approaches its population counterpart. The CLT describes the remaining estimation error and thereby makes confidence intervals and tests possible. Under the conditions used in this chapter, that error is approximately Gaussian in large samples even when the observations themselves are not.

**Prerequisites.** Expectations and variances, independent random variables, sample means, and convergence in probability.

**After working through this chapter, you should be able to:**

- Explain the distinct roles of the law of large numbers and the central limit theorem.
- Derive large-sample approximations for sums and sample means.
- Standardize a statistic and use normal quantiles to approximate probabilities.
- Compare means from two independent samples.

**Roadmap.** The chapter first establishes why sample averages converge, then studies their large-sample distribution. It concludes by applying these results to differences between independent sample means and to the exercises in Section \@ref(clt-exercises).


## Law of large numbers


:::{.definition #smplpop name="Sample and Population"}
In statistics, a *sample* refers to a finite set of observations drawn from a *population*.
:::

Empirical work usually relies on a sample because observing an entire population is impractical. A natural estimator of a population mean is the sample average of $\{x_1,\dots,x_n\}$:
$$
\overline{x}_n = \frac{1}{n}\sum_{i=1}^n x_i.
$$
This sample mean is an estimator of the true (population) mean $\mathbb{E}(x_i)$ (assuming the latter exists). According to the law of large numbers (Theorem \@ref(thm:LLN)), this estimator is *consistent*. (see Def. \@ref(def:convergenceproba) for the definition of convergence in probability.)

:::{.theorem #LLN name="Law of large numbers"}

If the random variables $x_i$ are i.i.d. and $\mathbb{E}(|x_i|)<\infty$, then the sample mean ($\overline{x}_n$) is a consistent estimator of the population mean $\mu=\mathbb{E}(x_i)$. In other words, the sample mean converges in probability to the population mean.
:::
:::{.proof}
See Appendix \@ref(CLTappend).
:::

:::{.example #doctorVisits}
Suppose one is interested in the average number of doctor consultations over 12 months, for Swiss people. One can get an estimate by computing the sample average of a large dataset, for instance using the [Swiss Household Panel (SHP)](https://forscenter.ch/projects/swiss-household-panel/):

Figure \@ref(fig:incomeDistri) shows the empirical distribution of the number of doctor visits.

```{r incomeDistri, fig.cap="Distribution of the number of doctor visits. Source: Swiss Household Panel.", fig.align = 'left'}
library(AEC) # to load the shp data-frame.
shp <- haven::zap_labels(AEC::shp) # retain values, remove import-only labels
Nb.doct.visits <- shp$p19c15
Nb.doct.visits <- Nb.doct.visits[Nb.doct.visits>=0] # remove irrelevant observations
par(plt=c(.15,.95,.2,.95))
barplot(table(Nb.doct.visits),xlim=c(0,25))
```
Let us compute the sample mean, variance, and standard deviation:

```{r incomeDistri2}
data_table(data.frame(Mean=mean(Nb.doct.visits),
                      Variance=var(Nb.doct.visits), SD=sd(Nb.doct.visits)), digits=3)
```

:::

Consistency tells us that the approximation improves with the sample size, but it does not yet quantify the error in a given sample. We first obtain a probability bound and then turn to the richer approximation supplied by the CLT.

A first answer is provided by the Chebychev inequality. Assume the $x_i$'s are independently drawn from a distribution of mean $\mu$ and variance $\sigma^2$. We have:
$$
\mathbb{E}(\bar{x}_n) = \mu \quad \mbox{ and } \quad \mathbb{V}ar(\mu - \bar{x}_n) = \frac{1}{n}\sigma^2 \underset{n \rightarrow \infty}{\rightarrow} 0.
$$
The Chebychev inequality (Proposition \@ref(prp:chebychev)) states that, for any $\varepsilon > 0$ (even very small), we have:
$$
\mathbb{P}(|\bar{x}_n-\mu|>\varepsilon) \le \frac{\mathbb{V}ar(\mu - \bar{x}_n)}{\varepsilon^2} = \frac{\sigma^2}{n\varepsilon^2} \underset{n \rightarrow \infty}{\rightarrow} 0.
$$
Consider for instance the estimation of the average number of doctor visits (Example \@ref(exm:doctorVisits)). In this example, the sample length is $n=$ `r length(Nb.doct.visits)`, we  have $\bar{x}_n=$ `r round(mean(Nb.doct.visits),2)`, and the variance $\sigma^2$ was approximately equal to `r round(var(Nb.doct.visits),2)`. Therefore, taking $\varepsilon=0.5$, we have that $\mathbb{P}(|\bar{x}_n-\mu|>\varepsilon)$ is lower than $\frac{\sigma^2}{n\varepsilon^2} \approx$
`r round(var(Nb.doct.visits)/(length(Nb.doct.visits)*0.5^2),3)`.

However, this only gives bounds for such probabilities. The central limit theorem provides richer information, as it gives the approximate distribution of the estimation error.

## Central Limit Theorem (CLT)

:::{.theorem #LindbergLevyCLT name="Lindberg-Levy Central limit theorem, CLT"}

If $x_n$ is an i.i.d. sequence of random variables with mean $\mu$ and (finite) variance $\sigma^2$, then:
$$
\boxed{\sqrt{n} (\bar{x}_n - \mu) \overset{d}{\rightarrow} \mathcal{N}(0,\sigma^2), \quad \mbox{where} \quad \bar{x}_n = \frac{1}{n} \sum_{i=1}^{n} x_i.}
$$
["$\overset{d}{\rightarrow}$" denotes the convergence in distribution (see Def. \@ref(def:cvgceDistri))]
:::
:::{.proof}
See Appendix \@ref(CLTappend).
:::

According to the CLT, when $n$ is large, the distribution of the error $\bar{x}_n-\mu$ is approximately normal, whatever the distribution of the $x_i$'s, provided that their mean $\mu$ and variance $\sigma^2$ exist. Its approximate mean is zero and its standard deviation is $\sigma/\sqrt{n}$. (We then say that the convergence rate is *in $\sqrt{n}$*.) Quantiles of the normal distribution can therefore be used to compute approximate confidence intervals for $\mu$ from the sample $\{x_1,\dots,x_n\}$.

This [web interface](https://jrenne.shinyapps.io/StatEcoII/) illustrates the CLT with simulations (select the "CLT" block).

The CL theorem was first postulated by the mathematician Abraham de Moivre. In an article published in 1733, he used the normal distribution to approximate the distribution of the number of heads resulting from many tosses of a fair coin. This finding was nearly forgotten until Pierre-Simon Laplace recalled it in his [Théorie analytique des probabilités](https://archive.org/details/thorieanalytiqu01laplgoog), published in 1812. Laplace expanded De Moivre's finding by approximating the binomial distribution with the normal distribution.


**Practice:** Exercises \@ref(exr:clt-buffon-precision) and \@ref(exr:clt-coin-normal-quantiles) use the CLT to design a simulation and to determine its required precision.

The CLT can be extended to the case where each $x_i$ is a vector. (In that case, and if the dimension of $x_i$ is $m$, then $\mathbb{V}ar(x_i)$ is a $m\times m$ matrix.) Here is the multivariate CLT:

:::{.theorem #MCLT name="Multivariate Central limit theorem, CLT"}
If $\bv{x}_n$ is an i.i.d. sequence of $m$-dimensional random variables with mean $\boldsymbol\mu$ and variance $\Sigma$, where $\Sigma$ is a positive definite matrix, then:
$$
\boxed{\sqrt{n} (\bar{\bv{x}}_n - \boldsymbol\mu) \overset{d}{\rightarrow} \mathcal{N}(\bv{0},\Sigma), \quad \mbox{where} \quad \bar{\bv{x}}_n = \frac{1}{n} \sum_{i=1}^{n} \bv{x}_i.}
$$
:::

## Comparison of sample means {#comparison-means}

The one-sample CLT extends directly to comparisons between populations. The key object is then the difference between two independent sample means.

Consider two independent samples $\{m_1,\dots,m_{n_m}\}$ and $\{w_1,\dots,w_{n_w}\}$. Assume that the $m_i$'s are i.i.d. with mean $\mu_m$ and variance $\sigma_m^2$, and that the $w_i$'s are i.i.d. with mean $\mu_w$ and variance $\sigma_w^2$.

If $n_m$ and $n_w$ are large, the CLT gives the approximations:
$$
\sqrt{n_m}(\bar{m} - \mu_m) \overset{a}{\sim} \mathcal{N}(0,\sigma^2_m) \quad \mbox{and} \quad \sqrt{n_w}(\bar{w} - \mu_w) \overset{a}{\sim} \mathcal{N}(0,\sigma^2_w),
$$
or
$$
\bar{m} \overset{a}{\sim} \mathcal{N}\left(\mu_m,\frac{\sigma^2_m}{n_m}\right) \quad \mbox{and} \quad \bar{w} \overset{a}{\sim} \mathcal{N}\left(\mu_w,\frac{\sigma^2_w}{n_w}\right).
$$

Since the two samples are independent, $\bar{m}$ and $\bar{w}$ are independent, and we get:
$$
\bar{m} - \bar{w} \overset{a}{\sim} \mathcal{N}\left(\mu_m-\mu_w,\tau^2\right) \quad \mbox{with} \quad \tau^2 = \frac{\sigma^2_m}{n_m}+\frac{\sigma^2_w}{n_w}.
$$
This can be used to test the null hypothesis $H_0: \; \mu_m-\mu_w = 0$ (see Section \@ref(Tests)). In practice, the unknown population variances are replaced by their sample counterparts, giving the estimated standard error $\hat\tau$. Under $H_0$, the probability that $\bar{m}-\bar{w}$ lies in $[-1.96\hat\tau,1.96\hat\tau]$ (respectively $[-2.58\hat\tau,2.58\hat\tau]$) is approximately 0.95 (respectively 0.99) in large samples [see Table \@ref(tab:Normal)].

For instance, if $\bar{m}-\bar{w}$ is not in $[-2.58\hat\tau,2.58\hat\tau]$, then we reject the null hypothesis at the 1\% significance level.

:::{.example #comparSmplMeans name="Comparison of sample means"}

Let us use the  [Swiss Household Panel (SHP)](https://forscenter.ch/projects/swiss-household-panel/) dataset to test the equality between men and women incomes, for Swiss people under the age of 35.

The SHP observations are split into two income samples, one for men and one for women. Observations with negative reported income are excluded.

```{r RcompMeans0}
library(AEC);library(logKDE)
dataset.m <- subset(shp,(sex19==1)&(age19<=35))
dataset.w <- subset(shp,(sex19==2)&(age19<=35))
income.m <- dataset.m$i19ptotg
income.w <- dataset.w$i19ptotg
income.m <- income.m[income.m>=0] # remove irrelevant data
income.w <- income.w[income.w>=0] # remove irrelevant data
```

Figure \@ref(fig:RcompMeans) displays the empirical densities of incomes for men and women. Vertical dashed lines indicate sample means:

```{r RcompMeans, fig.cap="Distribution of yearly gross incomes for Swiss residents under the age of 35. Source: SHP. Vertical dashed lines indicates the sample mean.", fig.align = 'left'}
par(plt=c(.1,.95,.2,.95)) # adjust margins
plot(logdensity(income.w),lwd=2,xlim=c(0,200000),main="",
     xlab="Yearly gross income (in CHF)",ylab="") # x and y labels
lines(logdensity(income.m),col="red",lwd=2)
abline(v=mean(income.m),col="red",lty=3,lwd=2)
abline(v=mean(income.w),col="black",lty=3,lwd=2)
legend("topright",c("men yearly income","women"),lty=1,lwd=2,col=c("red","black"))
```

Let us compute $\bar{m}$, $\bar{w}$, and the estimated standard error $\hat\tau$:

```{r RcompMeans2}
m.bar <- mean(income.m)
w.bar <- mean(income.w)
sigma <- sqrt(var(income.m)/length(income.m) + var(income.w)/length(income.w))
print(c(m.bar,w.bar,sigma))
```

Under the null hypothesis of equal population means, the difference $\bar{m}-\bar{w}$ would fall outside $[$`r -round(1.96*sigma,0)`,`r round(1.96*sigma,0)`$]$ with probability approximately 5\%, and outside $[$`r -round(2.58*sigma,0)`,`r round(2.58*sigma,0)`$]$ with probability approximately 1\%. The observed difference is `r format(round(m.bar-w.bar,0),scientific=FALSE)`, so we reject equality at both significance levels.
:::





<!-- %\begin{frame} -->
<!-- %\begin{footnotesize} -->
<!-- %		\begin{figure} -->
<!-- %		\caption{The importance of the independence assumption} -->
<!-- %			\includegraphics[width=1\linewidth]{../../figures/Figure_CLT_indpdce.pdf} -->
<!-- % -->
<!-- %		\begin{tiny} -->
<!-- %		\flushleft{Note: The three samples have been simulated using the following data generating process: $x_i = \mu + \rho (x_{i-1}-\mu) + \sqrt{1-\rho^2}\varepsilon_i$, where $\varepsilon_i \sim \mathcal{N}(0,1)$. Case A: $\rho = 0$;  Case B: $\rho = 0.9$;  Case C: $\rho = 0.99$. In the three cases, $\mathbb{E}(X_i)=\mu=2$ and $\mathbb{V}ar(X_i)=1$.} -->
<!-- %		\end{tiny} -->
<!-- %		\end{figure} -->
<!-- % -->
<!-- %\end{footnotesize} -->
<!-- %\end{frame} -->





<!-- %\begin{frame} -->
<!-- %\begin{footnotesize} -->
<!-- %\begin{exercise}[Polls and confidence intervals] -->
<!-- %\begin{itemize} -->
<!-- %	\item The day before an election with two candidates (A and B), 100 persons are surveyed (randomly chosen). 51\% of them say they will vote for Candidate A. -->
<!-- %	\item[(a)] Assuming people do not change their mind before the vote, compute an approximation of the probability that A will win. -->
<!-- %	 -->
<!-- %	(Hints: introduce $\theta$, the score that will be obtained by A; compute $\mathbb{P}(\theta > \mu + \bar{x}_n)$, where $x_i$ is equal to 1 if the $i^{th}$ respondent will votes for A and 0 otherwise). -->
<!-- %	\item[(b)] What is the probability if 55\% of them say they will vote for Candidate A. -->
<!-- %	\item[(c)] What would be that probability if 4000 people had been surveyed? (with 51\%) -->
<!-- %\end{itemize} -->
<!-- %\end{exercise} -->
<!-- %\begin{exercise}[Computer-free simulation of $\mathcal{N}$] -->
<!-- %\begin{itemize} -->
<!-- %	\item Assume you have a lot of time and a coin. How would you approximate the 100 percentiles of the Normal distribution? -->
<!-- %\end{itemize} -->
<!-- %\end{exercise} -->
<!-- %% Research on TCL and Casino: \url{http://digitalscholarship.unlv.edu/cgi/viewcontent.cgi?article=1293&context=grrj} -->
<!-- %\end{footnotesize} -->
<!-- %\end{frame} -->

```{r clt-recap-space, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) cat("\\Needspace{11\\baselineskip}\n")
```

## Chapter recap {#clt-recap}

::: {.recapbox data-latex=""}

- Theorem \@ref(thm:LLN) explains why a sample mean converges to its population mean; it is a consistency result.
- Chebyshev's inequality bounds the probability of a large estimation error but does not describe its full distribution.
- Theorem \@ref(thm:LindbergLevyCLT) gives the large-sample distribution of the standardized sample mean and therefore supports confidence intervals and tests.
- The multivariate extension in Theorem \@ref(thm:MCLT) applies the same logic to vectors of sample moments.
- Section \@ref(comparison-means) combines two independent CLT approximations to compare population means.

:::

```{r clt-exercise-pagebreak, echo=FALSE, results='asis'}
if (knitr::is_latex_output()) cat("\\clearpage\n")
```

## Exercises {#clt-exercises}

Attempt each question and explain your reasoning, including for true-or-false statements. **Foundation** exercises apply a result directly; **Standard** exercises require several steps or a short proof; **Advanced** exercises combine several results or require a longer derivation. Worked solutions and exercise-specific hints are reserved for teaching sessions and are not included in this student edition.

```{r clt-exercise-bank, echo=FALSE, results='asis'}
source("R/exercises.R")
render_exercises("clt")
```
