--- 
title: "Econometrics and Statistics"

# ======================================================================
# SHOW EXERCISE SOLUTIONS? Change only false / true below.
# false = STUDENT VERSION: exercises only; no hints or solutions.
# true  = PRIVATE INSTRUCTOR VERSION: hints and solutions after exercises.
# Applies to BOTH PDF and HTML. Use the usual scripts/build*.R command.
# Instructor output is saved outside this repository in Instructor_materials.
# Set back to false before building/publishing the student version.
# ======================================================================
params:
  show_solutions: false
# ======================================================================

# ======================================================================
# SHOW R CODE IN THE PDF? Change only false / true below.
# false = code is hidden in the PDF; figures, tables, and results remain.
# true  = code is shown in the PDF.
# This switch affects ONLY the PDF. Code remains visible in the HTML book.
# ======================================================================
  show_code_in_pdf: false
# ======================================================================

author: "Jean-Paul Renne"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
fontsize: 11pt
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: jrenne/EcoStat
description: "A textbook introduction to probability, statistics, econometrics, and time-series analysis, with simulations, applications, and exercises."
css: "my-style.css"
---

```{r table-setup, include=FALSE}
source("R/instructor-build.R")
configure_edition(params$show_solutions)
source("R/setup.R")
configure_code_display(params$show_code_in_pdf)
```


<!-- Declare new latex code: -->
\newcommand{\bv}[1]{\mathbf{#1}} 

# Econometrics and statistics {#intro .unlisted .unnumbered}

<!-- # Before starting {#intro .unlisted .unnumbered} -->

This book develops the statistical foundations of econometrics and then uses them to study regression, panel data, parametric estimation, binary-choice models, and time series. Mathematical results are paired with simulations, empirical applications, and exercises. The material has been developed by [Jean-Paul Renne](https://www.jprenne.com).

The chapters are cumulative. The first three establish the probability and inference tools used later. The linear-regression chapter develops the central econometric framework; the chapters on panel regressions, estimation methods, and binary-choice models build on it. The time-series chapter introduces time-indexed data and forecasting. The appendix collects supporting results and longer proofs so that the main chapters can retain their narrative flow.

Definitions and propositions state the core results. Examples show how those results operate in data, while practice boxes point to exercises that reinforce the surrounding material. Readers who use the computational examples can inspect the R code in the HTML edition; the PDF can be rendered with or without that code.

The computational material uses various packages available from [CRAN](https://cran.r-project.org), as well as procedures and data from the companion package `AEC`. This package is available on GitHub. Installation instructions and the recorded package environment are described in the repository README.

```{r installAEC, eval=FALSE}
install.packages("remotes")
remotes::install_github("jrenne/AEC@3245789138e77c2d7b591b40e8b6a3405e93cbf1")
library(AEC)
```

**Useful R links:**

* Download R:

    *	R software: https://cran.r-project.org (the basic R software)
    *	RStudio: https://www.rstudio.com (a convenient R editor)

* Tutorials:

    *	Rstudio: https://dss.princeton.edu/training/RStudio101.pdf (by Oscar Torres-Reyna)
    *	R: https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf (by Emmanuel Paradis)
    * My own tutorial: https://jrenne.shinyapps.io/Rtuto_publiShiny/
